---
title: "Spatial Block Tile Generation"
author: "Mia Huebner"
format: html
editor: visual
---

## Spatial Block Tile Generation

(Spatial) Block cross validation will be performed to assess how well the model fits the data. For this assessment, however, I need to create the spatial block tiles. I was able to do this through CRAN's documentation on the blockCV package (Valavi et al., 2024).

Load in the necessary libraries:

```{r}
library(blockCV)
library(sf)
library(terra)
library(geodata)
```

Load in the raster data of the Aquilegia distribution:

```{r}
C_Data <- "https://raw.githubusercontent.com/MiaHuebner/Aquilegia-Distribution-Final-Project/refs/heads/main/Use_This_Data_For_Block_Generation.txt"

Tile_Generation_Data <- read.table(C_Data, header = TRUE, sep = "\t")
Tile_Generation_Data_df <- data.frame(Tile_Generation_Data)
```

Change the data into a spatial object:

```{r}
Aquilegia_species_data <- sf::st_as_sf(Tile_Generation_Data_df, coords = c("decimalLongitude", "decimalLatitude"), crs = 4326)
```

Add the environmental predictors as a raster, this code is taken from my SDM analysis:

```{r}
Global_Current_Data <- worldclim_global(
  var = 'bio', 
  res = 5, 
  path = 'C:/Users/Mia/OneDrive/Desktop/RWorkPLUS/GISR/FinalProject/AquilegiaFormosaDistribFinalProj/CurrentWorldClim', #Change this line of code to download the data onto your own computer 
  download = FALSE)[[c(2, 5, 14)]]

summary(Global_Current_Data)

North_America_Boundary <- geodata::gadm(
  country = c("USA", "Canada", "Mexico"), 
  level = 0, 
  path = tempdir('C:/Users/Mia/OneDrive/Desktop/RWorkPLUS/GISR/FinalProject/AquilegiaFormosaDistribFinalProj')
)

raster_of_the_three <- rast(
  ext(North_America_Boundary), 
  resolution = 0.5, 
  crs = crs(North_America_Boundary)
)

mask <- rasterize(North_America_Boundary, raster_of_the_three)
NA_coords <- ext(-175, -50, 5, 90)
mask <- crop(mask, NA_coords)

Aquilegia_Range_Longitude <- -155:-108
Aquilegia_Range_Latitude <- 29:66
Aquilegia_Range <- terra::ext(min(Aquilegia_Range_Longitude), max(Aquilegia_Range_Longitude),
                          min(Aquilegia_Range_Latitude), max(Aquilegia_Range_Latitude))

Current_Bio <- crop(Global_Current_Data, Aquilegia_Range)
Current_Bio <- project(Current_Bio, mask)
Current_Bio <- resample(Current_Bio, mask)
Current_Bio <- mask(Current_Bio, mask)
names(Current_Bio) <- c('Current.Mean.Diurnal.Range..Bio.2.', 'Current.Max..Temp..of.Warmest.Month..Bio.5.', 
                        'Current.Driest.Month.Precipitation.Data..Bio.14.')

Current_Bio[[1]] <- Current_Bio[[1]]/10
Current_Bio[[2]] <- Current_Bio[[2]]/10


```

Now, define the spatial blocks, I used 420 kilometers as the block range because it made 8 blocks fit into the mask frame well:

```{r}
spatial_blocks <- blockCV::spatialBlock(
  speciesData = Aquilegia_species_data, 
  rasterLayer = Current_Bio, 
  theRange = 420000,
  k = 8,
  selection = "random", 
  iteration = 50, 
  biomod2Format = TRUE, 
  xOffset = 0, 
  yOffset =0
)
```

It says that the function is deprecated, and to use cv_spatial instead, however, the blocks are still generated well, so I ignored this message.

Now, add fold assignments to the Aquilegia data:

```{r}
Aquilegia_species_data$blockCV_tile <- spatial_blocks$blockCV_tile
```

Change the data back into a dataframe, then save it as a CSV file:

```{r}
Aquilegia_species_df <- data.frame(Aquilegia_species_data)
Aquilegia_species_df

write.csv(Aquilegia_species_data, "BlockCVInfo", row.names = TRUE)
```

Generating the spatial blocks was a lot less painful than generating the pseudo-absence data! After the spatial block generation, I used excel to double check the file and make sure all necessary data was present.

Again, I was able to generate this data through the blockCV package documentation available on CRAN at https://cran.r-project.org/web/packages/blockCV/vignettes/tutorial_2.html, but from Valavi et al. 2019:

Valavi R, Elith J, Lahoz-Monfort JJ, Guillera-Arroita G (2019). "blockCV: An R package for generating spatially or environmentally separated folds for k-fold cross-validation of species distribution models." *Methods in Ecology and Evolution*, **10**(2), 225-232. [doi:10.1111/2041-210X.13107](https://doi.org/10.1111/2041-210X.13107)
